Figures_MSA-PD

### 2025-08-22 Manuscript_figures.Rmd
Update for revision

```{r setup, message = F}
library(conos)
library(magrittr)
library(dplyr)
library(cacoa) # github.com/kharchenkolab/cacoa
library(sccore)
library(scHelper) # github.com/rrydbirk/scHelper
library(qs)
library(ggplot2)
library(ggsci)
library(cowplot)
library(ggpubr)
library(STRINGdb)
library(reshape2)
library(ggforce)
library(ggpmisc)
library(RColorBrewer)
library(CRMetrics)
library(ggrepel)
library(circlize)
library(ComplexHeatmap)
library(stringr)
library(rstatix)
library(slingshot)

@@

```{r}
getOntWithFamily <- function(cao, comp) {
  fams <- cao$test.results[["GSEA"]]$families

  df.all <- list(cao$.__enclos_env__$private$getOntologyPvalueResults(name = "GSEA", genes = "up", p.adj = 1, q.value = 1),
                 cao$.__enclos_env__$private$getOntologyPvalueResults(name = "GSEA", genes = "down", p.adj = 1, q.value = 1)) %>% 
    bind_rows() %>% 
    mutate(logp = -log10(p.adjust),
           comp = comp)

  df <- cacoa:::getOntologyFamilyChildren(df.all, fams=fams, type = "GSEA", subtype = "BP")

@@

```{r, fig.width=5.7, fig.height=4}
con$plotGraph(groups = anno.major, 
              embedding = "UMAP",
              size = 0.1, 
              palette = pal.major, 
              font.size = 3, 
              raster = T, 
              show.labels = T,
              plot.na = F, 
              show.legend = T,
              legend.title = "Cell type", 
              alpha = 0.05) + 
  dotSize(3) +
  labs(x="UMAP1", y= "UMAP2") +
  theme(line = element_blank())
```

@@

@@

con$plotGraph(groups = cpc, 
              embedding = "UMAP",
              size = 0.1, 
              alpha = 0.05,
              palette = pal.major, 
              font.size = 3, 
              raster = T, 
              show.labels = T,
              plot.na = F,
              mark.groups = F,
              show.legend = T) + 
  labs(x="UMAP1", y= "UMAP2", col = "") +
  theme(legend.position = "bottom",
        line = element_blank()) +
  dotSize(3)
```

@@

```{r, fig.width = 6, fig.height = 4}
cao_msa$cell.groups.palette <- pal.major
cao_pd$cell.groups.palette <- pal.major
cao_dis$cell.groups.palette <- pal.major

cao_msa$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
cao_pd$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
cao_dis$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
```

@@

@@

cao$plotCellDensity(show.cell.groups = F)$CTRL +
  geom_circle(aes(x0 = 0.75, y0 = -4.5, r = 1)) +
  geom_circle(aes(x0 = 3.8, y0 = -1.1, r = 0.5), col = "cyan3") +
  theme(line = element_blank())

# MSA
cao$sample.groups <- sample.groups %>% 
  names() %>% 
  grepl.replace(c("CTRL","MSA","PD")) %>% 
  `names<-`(sample.groups %>% names()) %>% 
  .[. %in% c("CTRL","MSA")]
cao$target.level <- "MSA"
cao$estimateCellDensity(method = "graph", name = "msa.density")
cao$estimateDiffCellDensity(type = "subtract", name = "msa.density")
cao$plotCellDensity(show.cell.groups = F, name = "msa.density")$MSA +
  geom_circle(aes(x0 = 0.75, y0 = -4.5, r = 1)) +
  geom_circle(aes(x0 = 3.8, y0 = -1.1, r = 0.5), col = "cyan3") +
  theme(line = element_blank())

# PD
cao$sample.groups <- sample.groups %>%
  names() %>%
  grepl.replace(c("CTRL","MSA","PD")) %>%
  `names<-`(sample.groups %>% names()) %>%
  .[. %in% c("CTRL","PD")]
cao$target.level <- "PD"
cao$estimateCellDensity(method = "graph", name = "pd.density")
cao$estimateDiffCellDensity(type = "subtract", name = "pd.density")
cao$plotCellDensity(show.cell.groups = F, name = "pd.density")$PD +
  geom_circle(aes(x0 = 0.75, y0 = -4.5, r = 1)) +
  geom_circle(aes(x0 = 3.8, y0 = -1.1, r = 0.5), col = "cyan3") +
  theme(line = element_blank())
```

## Figure 3g

@@

## Figure 4a

@@

## Figure 4b

@@

## Figure 4d

@@

## Figure 4e

@@

## Figure 4h,i

@@

### Figure 4h

@@

### Figure 4i

@@

## Figure 4j

@@

## Figure 4k,l

@@

### Figure 4k

@@

### Figure 4l

@@

## Figure 4m

@@

# Figure 5 & EDF7

@@

## Figure 5d-f

@@

@@

## Figures 5e,f & EDF7

@@

```{r, eval = F}
library(gamm4)

res <- mt %>% 
  sccore::plapply(\(x) {
    fit_full <- gamm4(data = x, formula = value ~ s(pseudotime, by = factor(condition)), random = ~(1 | sample), REML = F)
    fit_reduced <- gamm4(data = x, formula = value ~ s(pseudotime), random = ~(1 | sample), REML = F)
    fit_none <- gamm4(data = x, formula = value ~ 1, random = ~(1 | sample), REML = F)

@@

### Figure 5f

@@

### Extended Data Figure 7

@@

@@

## Figure 5i

@@

@@

## Figure 5m

@@

@@

@@

@@

cao$estimateExpressionShiftMagnitudes()
cao$estimateMetadataSeparation(sample.meta = meta, 
                               space = "expression.shifts")
```

@@

@@

## 3c

@@

## 3d

@@

# Extended Data Figure 5

>>>>

```{r setup, message = F}
library(conos)
library(magrittr)
library(dplyr)
library(cacoa) # github.com/kharchenkolab/cacoa
library(sccore)
library(scHelper) # github.com/rrydbirk/scHelper
library(qs)
library(ggplot2)
library(ggsci)
library(cowplot)
library(ggpubr)
library(STRINGdb)
library(reshape2)
library(ggforce)
library(ggpmisc)
library(RColorBrewer)
library(CRMetrics)
library(ggrepel)
library(circlize)
library(ComplexHeatmap)
library(stringr)
library(rstatix)
library(slingshot)
library(gamm4)
library(scITD)
library(ggraph)

@@

```{r}
getOntWithFamily <- function(cao, comp, name = "GSEA") {
  fams <- cao$test.results[[name]]$families

  df.all <- list(cao$.__enclos_env__$private$getOntologyPvalueResults(name = name, genes = "up", p.adj = 1, q.value = 1),
                 cao$.__enclos_env__$private$getOntologyPvalueResults(name = name, genes = "down", p.adj = 1, q.value = 1)) %>% 
    bind_rows() %>% 
    mutate(logp = -log10(p.adjust),
           comp = comp)

  df <- cacoa:::getOntologyFamilyChildren(df.all, fams=fams, type = name, subtype = "BP")

@@

```{r, fig.width=5.7, fig.height=4}
con$plotGraph(groups = anno.major, 
              embedding = "UMAP_1_0.001_5",
              size = 0.1, 
              palette = pal.major, 
              font.size = 3, 
              raster = T, 
              show.labels = T,
              plot.na = F, 
              show.legend = T,
              legend.title = "Cell type", 
              alpha = 0.05) + 
  dotSize(3) +
  labs(x="UMAP1", y= "UMAP2") +
  theme(line = element_blank()) +
  ylim(c(-20, 15)) +
  xlim(c(-21, 11))
```

@@

con$plotGraph(groups = cpc, 
              embedding = "UMAP_1_0.001_5",
              size = 0.1, 
              alpha = 0.05,
              palette = pal.major, 
              font.size = 3, 
              raster = F, 
              show.labels = T,
              plot.na = F,
              mark.groups = F,
              show.legend = T) + 
  labs(x="UMAP1", y= "UMAP2", col = "") +
  theme(legend.position = "bottom",
        line = element_blank()) +
  dotSize(3)
```

## Figure 1e

## Figure 1f

## Figure 1g

## Figure 1h

```{r, fig.width = 6, fig.height = 4}
cao_msa$cell.groups.palette <- pal.major
cao_msa$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
```

## Figure 1i

```{r}
cao_pd$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
cao_pd$cell.groups.palette <- pal.major
```

## Figure 1j

```{r}
cao_dis$plotExpressionShiftMagnitudes() + labs(y = "Normalized expression distance")
cao_dis$cell.groups.palette <- pal.major
```

@@

@@

@@

## Figure 2f

## Figure 2g

@@

cao$estimateCellDensity(method = "graph")
cao$estimateDiffCellDensity(type = "subtract")

cao$plotCellDensity(show.cell.groups = F, show.legend = F, color.range = c(0, 0.00035))$CTRL +
  geom_circle(aes(x0 = 30.9, y0 = 10, r = 10)) +
  geom_circle(aes(x0 = 17.5, y0 = 24, r = 4), col = "cyan3") +
  theme(line = element_blank())
```

```{r, fig.height=4, fig.width=4}
# MSA
cao$sample.groups <- sample.groups %>% 
  names() %>% 
  grepl.replace(c("CTRL","MSA","PD")) %>% 
  `names<-`(sample.groups %>% names()) %>% 
  .[. %in% c("CTRL","MSA")]
cao$target.level <- "MSA"
cao$estimateCellDensity(method = "graph", name = "msa.density")
cao$estimateDiffCellDensity(type = "subtract", name = "msa.density")
cao$plotCellDensity(show.cell.groups = F, name = "msa.density", show.legend = F, color.range = c(0, 0.00035))$MSA +
  geom_circle(aes(x0 = 30.9, y0 = 10, r = 10)) +
  geom_circle(aes(x0 = 17.5, y0 = 24, r = 4), col = "cyan3") +
  theme(line = element_blank())
```

```{r, fig.height=4, fig.width=5.1}
# PD
cao$sample.groups <- sample.groups %>%
  names() %>%
  grepl.replace(c("CTRL","MSA","PD")) %>%
  `names<-`(sample.groups %>% names()) %>%
  .[. %in% c("CTRL","PD")]
cao$target.level <- "PD"
cao$estimateCellDensity(method = "graph", name = "pd.density")
cao$estimateDiffCellDensity(type = "subtract", name = "pd.density")
cao$plotCellDensity(show.cell.groups = F, name = "pd.density", show.legend = T, color.range = c(0, 0.00035))$PD +
  geom_circle(aes(x0 = 30.9, y0 = 10, r = 10)) +
  geom_circle(aes(x0 = 17.5, y0 = 24, r = 4), col = "cyan3") +
  theme(line = element_blank())
```

@@

@@

@@

@@

@@

```{r, echo=F}
rm(con.glia)
gc()
```

## Figure 4a,b

@@

### Figure 4a

@@

### Figure 4b

@@

@@

## Figure 4d,e

@@

### Figure 4d

@@

### Figure 4e

@@

@@

# Figure 5 & EDF13

@@

## Figure 5c-e

@@

### Figure 5c

@@

## Figures 5d,e & EDF13

@@

```{r, eval = F}
res <- mt %>% 
  sccore::plapply(\(x) {
    fit_full <- gamm4(data = x, formula = value ~ s(pseudotime, by = factor(condition)), random = ~(1 | sample), REML = F)
    fit_reduced <- gamm4(data = x, formula = value ~ s(pseudotime), random = ~(1 | sample), REML = F)
    fit_none <- gamm4(data = x, formula = value ~ 1, random = ~(1 | sample), REML = F)

@@

@@

```{r, echo = F}
rm(cao)
gc()
```

### Extended Data Figure 13

@@

## Figure 5f

@@

@@

```{r, echo = F}
rm(cao_dis)
gc()
```

## Figure 5j

@@

```{r, echo = F}
rm(cao_msa)
gc()
```

@@

```{r, echo = F}
rm(cao_pd)
gc()
```

@@

## Figure 6j

Here, we use data from [Rydbirk *et al*.](https://pmc.ncbi.nlm.nih.gov/articles/PMC9164190/) on differentially expressed proteins in the soluble fraction of CSF pools.

```{r, fig.width=6, fig.height=4}
dat <- read.delim("Soluble_fraction.txt")

msa.dep <- dat %>% 
  filter(Disease.group == "MSA") %>% 
  pull(Gene.name)

pd.dep <- dat %>% 
  filter(Disease.group == "PD") %>% 
  pull(Gene.name)

overlap <- intersect(msa.dep, pd.dep)

highlight_proteins <- c("CTSS", "FCGR2A", "LAMP1", "CTSD", "CLTC")

msa.dep %<>% .[!. %in% overlap]
pd.dep %<>% .[!. %in% overlap]
overlap %<>% .[!. %in% highlight_proteins]

string_db <- STRINGdb$new(
  version = "12.0",       # or latest available
  species = 9606,         # 9606 = Homo sapiens
  score_threshold = 400,  # confidence score cutoff
  input_directory = ""
)

genes <- c(msa.dep, pd.dep, overlap)

mapped_genes <- string_db$map(
  data.frame(gene = genes),
  "gene",
  removeUnmappedRows = TRUE
)

# Get interactions
interactions <- string_db$get_interactions(mapped_genes$STRING_id)

# Convert to igraph object
ppi_graph <- graph_from_data_frame(interactions, directed = FALSE)

# Create label mapping: STRING ID to gene name
string_to_gene <- mapped_genes %>%
  select(STRING_id, gene) %>%
  distinct()

# Add labels to graph nodes
V(ppi_graph)$label <- string_to_gene$gene[match(V(ppi_graph)$name, string_to_gene$STRING_id)]

custom_colors <- c(
  "CTSS" = "firebrick",
  "FCGR2A" = "purple2",
  "LAMP1" = "purple2",
  "CTSD" = "navyblue",
  "CLTC" = "navyblue",
  setNames(rep("tomato", length(msa.dep)), msa.dep),
  setNames(rep("steelblue", length(pd.dep)), pd.dep),
  setNames(rep("orchid", length(overlap)), overlap))

# Apply color mapping to node attributes
V(ppi_graph)$node_color <- custom_colors[V(ppi_graph)$label]

# Get node labels for each edge endpoint
edge_ends <- ends(ppi_graph, es = E(ppi_graph), names = FALSE)
source_labels <- V(ppi_graph)$label[edge_ends[, 1]]
target_labels <- V(ppi_graph)$label[edge_ends[, 2]]

# Assign edge width
E(ppi_graph)$edge_width <- ifelse(
  source_labels %in% highlight_proteins & target_labels %in% highlight_proteins,
  0.5, 0.1
)

# Compute node sizes based on label
V(ppi_graph)$node_size <- sapply(V(ppi_graph)$label, \(x) if (x %in% overlap) 3 else if (x %in% highlight_proteins) 5 else 2)

V(ppi_graph)$label_type <- ifelse(
  V(ppi_graph)$label %in% highlight_proteins,
  "highlight", "other"
)

# Plot the network with size and color
ggraph(ppi_graph, layout = "fr") +
  geom_edge_link(aes(width = I(edge_width)), alpha = 0.4) +
  geom_node_point(aes(color = I(node_color), size = I(node_size)), show.legend = FALSE) +
  geom_node_label(data = function(x) { x[x$label_type == "highlight", ] },
                  aes(label = label), size = 3, label.padding = unit(0.15, "lines"), repel = TRUE, show.legend = F) +
  geom_node_text(data = function(x) { x[x$label_type == "other", ] },
                 aes(label = label), size = 2, repel = TRUE, show.legend = F) +
  theme_void()
```

@@

```{r, echo=FALSE}
rm(crm)
gc()
```

## 2c

@@

cao$estimateExpressionShiftMagnitudes()
cao$estimateMetadataSeparation(sample.meta = meta, 
                               space = "expression.shifts")

meta %<>% 
  tibble::rownames_to_column("sample")
```

@@

```{r, fig.width=18, fig.height=16}
spc <- con$getDatasetPerCell()

mpc <- spc %>% 
  data.frame(sample = ., cid = names(.))

for (cc in colnames(meta)[-1]) {
  mpc[[cc]] <- meta[[cc]][match(mpc$sample, meta$sample)]
}

mpc$subtype[mpc$subtype == ""] <- NA
```

```{r, fig.width=5, fig.height=4}
p <- con$plotGraph(colors = setNames(mpc$age, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, title = "Age", size = 0.3, alpha = 0.3) +
        scale_color_gradient(low = "grey", high = "firebrick") +
        theme(line = element_blank()) +
  labs(col = "Years")

p
```

```{r, fig.width=5, fig.height=4}
p <- con$plotGraph(colors = setNames(mpc$pmi, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, title = "PMI", size = 0.3, alpha = 0.3) +
        scale_color_gradient(low = "grey", high = "firebrick") +
        theme(line = element_blank()) +
  labs(col = "Hours")

p
```

```{r, fig.width=5, fig.height=4}
p <- con$plotGraph(colors = setNames(mpc$disease_duration, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, title = "Disease duration", size = 0.3, alpha = 0.3) +
        scale_color_gradient(low = "grey", high = "firebrick") +
        theme(line = element_blank()) +
  labs(col = "Years")

p
```

```{r, fig.width=6.75, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$sample, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Sample", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

```{r, fig.width=5.25, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$brain_bank, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Brain bank", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

```{r, fig.width=5.15, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$sex, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Sex", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

```{r, fig.width=5.4, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$subtype, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Subtype", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

```{r, fig.width=5.15, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$extract, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Extract", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

```{r, fig.width=5.88, fig.height=4}
p <- con$plotGraph(groups = setNames(mpc$flowcell, mpc$cid), plot.na = F, embedding = "UMAP_1_0.001_5", mark.groups = F, show.legend = T, shuffle.colors = T, title = "Flowcell", size = 0.3, alpha = 0.3) +
        theme(line = element_blank()) + dotSize(3)

p
```

@@

@@

@@

# Extended Data Figure 4

To investigate effects from co-variates, we ran scITD on all cell types for each comparison. We don't determine factors here as it takes a significant amount of time, but we keep the code in the first instance.

We tested different levels of **var_scale_power**, here we're just showing **var_scale_power = 0.5** as we didn't find any effects except for OPCs. Per default, we set **vargenes_method="norm_var_pvals"** and adjusted **vargenes_thresh** to obtain around 1,000 var. genes, but in some instances we took the top most variable genes instead.

### Neurons, preparation

```{r}
con.tmp <- qread("cao_neurons.qs", nthreads = 10)$data.object
anno <- qread("anno_neurons.qs")

anno %<>% 
  collapseAnnotation("GABA") %>%
  collapseAnnotation("GLU") %>%
  renameAnnotation("GABA", "Inh. neurons") %>%
  renameAnnotation("GLU", "Exc. neurons") %>%
  collapseAnnotation("MSN")

cm.raw.all <- con.tmp$getJointCountMatrix(raw = T) %>% 
  Matrix::t() %>% 
  .[!grepl("MT-|RPS|RPL", rownames(.)), ]

meta.all <- con.tmp$getDatasetPerCell() %>% 
  data.frame(donors = .) %>% 
  mutate(ctypes = anno[rownames(.)] %>% unname()) %>% 
  .[complete.cases(.), ]

metadata.all <- read.delim("metadata.tsv")

for (cc in colnames(metadata.all)[-1]) {
  meta.all[[cc]] <- metadata.all[[cc]][match(meta.all$donors, metadata.all$sample)]
}

# Convert integers to numeric
meta.all$age %<>% as.numeric()
meta.all$disease_duration %<>% as.numeric()
```

### CTRL vs MSA

```{r}
meta <- meta.all %>% 
  filter(condition != "PD") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "PD")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw %>% .[, colnames(.) %in% rownames(meta)], 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=1,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var_pvals', 
                               vargenes_thresh=0.7,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r, eval = F}
# get assistance with rank determination
itd_container %<>% determine_ranks_tucker(max_ranks_test=c(10,15),
                                          shuffle_level='cells',
                                          num_iter=10,
                                          norm_method='trim',
                                          scale_factor=10000,
                                          scale_var=TRUE,
                                          var_scale_power=0.5)

itd_container$plots$rank_determination_plot
```

```{r, fig.height=5, fig.width=8}
itd_container %<>% run_tucker_ica(ranks=c(5,10),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "flowcell", "extract"), # Omitting brain bank
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "flowcell", "extract"), # Omitting brain bank
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### CTRL vs PD

```{r}
meta <- meta.all %>% 
  filter(condition != "MSA") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "MSA")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw, 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=1,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var',
                               vargenes_thresh=500,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(5,10),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"),
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"),
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### PD vs MSA

```{r}
meta <- meta.all %>% 
  filter(condition != "CTRL") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "CTRL")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw, 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=1,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var', 
                               vargenes_thresh=500,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(5,10),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"),
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"),
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### Astrocytes, preparation

```{r}
con.tmp <- qread("con_astrocytes.qs", nthreads = 10)
anno <- qread("anno_astro.qs")

cm.raw.all <- con.tmp$getJointCountMatrix(raw = T) %>%
  .[, !grepl("MT-|RPS|RPL", colnames(.))] %>% 
  Matrix::t()

meta.all <- con.tmp$getDatasetPerCell() %>% 
  .[names(.) %in% names(anno)] %>% 
  data.frame(donors = .) %>% 
  mutate(ctypes = anno[rownames(.)] %>% unname()) %>% 
  .[complete.cases(.), ]

metadata.all <- read.delim("metadata.tsv")

for (cc in colnames(metadata.all)[-1]) {
  meta.all[[cc]] <- metadata.all[[cc]][match(meta.all$donors, metadata.all$sample)]
}

# Convert integers to numeric
meta.all$age %<>% as.numeric()
meta.all$disease_duration %<>% as.numeric()
```

### CTRL vs MSA

```{r}
meta <- meta.all %>% 
  filter(condition != "PD") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "PD")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw %>% .[, colnames(.) %in% rownames(meta)], 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=5,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var_pvals', 
                               vargenes_thresh=0.05,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(5,8),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "flowcell", "extract"), # Omitting brain bank
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "flowcell", "extract"), # Omitting brain bank
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### CTRL vs PD

```{r}
meta <- meta.all %>% 
  filter(condition != "MSA") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "MSA")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw, 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=5,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var_pvals', 
                               vargenes_thresh=.05,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(5,9),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"), # Omitting brain bank
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"), # Omitting brain bank
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### PD vs MSA

```{r}
meta <- meta.all %>% 
  filter(condition != "CTRL") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "CTRL")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw, 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=5,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var_pvals', 
                               vargenes_thresh=.05,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(6,7),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"),
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars=c('sex','pmi', "age", "brain_bank", "flowcell", "extract"), 
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### Microglia, preparation

```{r}
con.tmp <- qread("cao_micro_pvm.qs", nthreads = 10)$data.object
anno <- qread("anno_micro_pvm.qs")

cm.raw.all <- con.tmp$getJointCountMatrix(raw = T) %>% 
  Matrix::t() %>% 
  .[!grepl("MT-|RPS|RPL", rownames(.)), ]

meta.all <- con.tmp$getDatasetPerCell() %>% 
  data.frame(donors = .) %>% 
  mutate(ctypes = anno[rownames(.)] %>% unname()) %>% 
  .[complete.cases(.), ]

metadata.all <- read.delim("metadata.tsv")

for (cc in colnames(metadata.all)[-1]) {
  meta.all[[cc]] <- metadata.all[[cc]][match(meta.all$donors, metadata.all$sample)]
}

# Convert integers to numeric
meta.all$age %<>% as.numeric()
meta.all$disease_duration %<>% as.numeric()
```

### CTRL vs MSA

```{r}
meta <- meta.all %>% 
  filter(condition != "PD") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "PD")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw %>% .[, colnames(.) %in% rownames(meta)], 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=3,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var', 
                               vargenes_thresh=500,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```

```{r}
itd_container %<>% run_tucker_ica(ranks=c(3,7),
                                  tucker_type = 'regular', 
                                  rotation_type = 'hybrid')

factors <- c('sex','pmi', "age", "flowcell", "extract")

# get donor scores-metadata associations
itd_container %<>% get_meta_associations(vars_test = factors, 
                                         stat_use='pval')

# plot donor scores
itd_container %<>% plot_donor_matrix(meta_vars= factors,
                                     show_donor_ids = TRUE,
                                     add_meta_associations='pval')

# show the donor scores heatmap
p <- itd_container$plots$donor_matrix

p
```

### CTRL vs PD

```{r}
meta <- meta.all %>% 
  filter(condition != "MSA") %>% 
  mutate(donors = factor(donors),
         ctypes = factor(ctypes))

metadata <- metadata.all %>% 
  filter(condition != "MSA")

cm.raw <- cm.raw.all %>% 
  .[, colnames(.) %in% rownames(meta)]

# set up project parameters
param_list <- initialize_params(ctypes_use = meta$ctypes %>% levels(),
                                ncores = 31, 
                                rand_seed = 10)

# create project container
itd_container <- make_new_container(count_data=cm.raw, 
                                    meta_data=meta,
                                    params=param_list,
                                    label_donor_sex = F)

itd_container %<>% form_tensor(donor_min_cells=3,
                               norm_method='trim', 
                               scale_factor=10000,
                               vargenes_method='norm_var', 
                               vargenes_thresh=500,
                               scale_var = TRUE, 
                               var_scale_power = 0.5)

print(length(itd_container[["all_vargenes"]]))
```
